A type-safe generalized editor calculus (Short Paper)

### 2023-07-02 Article.tex
Ny indledning
Ny indledning.

\documentclass[sigplan,screen]{acmart}
\input{preamble}
%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[GPCE 2023]{Make sure to enter the correct
  conference title from your rights confirmation emai}{July 07--07,
  2023}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A generalised editor calculus (Short Paper)}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
In this paper, we present a generalization of a syntax-directed editor calculus, which can be used to instantiate a specialized syntax-directed editor for any language, given by some abstract syntax. The editor calculus guarantees the absence of syntactical errors while allowing incomplete programs. The generalized editor calculus is then encoded into a simply typed lambda calculus, extended with pairs, booleans, pattern matching and fixed points.
\end{abstract}

@@

\section{Introduction}
This paper is on the topic of syntax-directed editors. 
Most current examples of syntax-directed editor calculi, such as the ones presented by \cite{hazelnut} and \cite{type_safe_structure_editor}, are typically tailored specifically towards editing abstract syntax trees from an applied $\lambda$-calculus. 
This fact alone limits the usable scope of these calculi, since they would be rendered useless given any other form of abstract syntax.

Furthermore, these calculi do not account for binding mechanisms in abstract syntax. Many of the editor calculi possess the ability to interleave construction and evaluation of programs, however, a consequence of this is a complex type system.

In this paper we propose a generalized editor calculus that can be used to create a syntax-directed editor calculus for any abstract syntax. 
Our editor calculus is inspired by higher-order abstract syntax presented in \cite{hoas}, and it will take binding-mechanisms into account. We have encoded our generalized editor calculus into a simply-typed lambda calculus extended with pairs, booleans, pattern matching and fixed-points.

%Copy paste fra her
\section{Generalized Editor Calculus}\label{sec:general_editor}
We will now present a generalization of the syntax-directed editor calculus by \cite{type_safe_structure_editor}. 
This specific editor calculus was designed with the language of an applied $\lambda$-calculus in mind. 
%Instead, by using the notion of abstract syntax as presented in \cite{sec:pre_abstract_syntax} we generalize the concepts of the editor calculus to account for any language given by some abstract syntax.

\subsection{Syntax}

\subsubsection{Abstract Syntax}
Firstly, we introduce the general representation of some abstract syntax that will be the focus of the editor calculus. We assume the abstract syntax is given by a set of sorts $\SORT$, an arity-indexed family of operators $\OP$ and a sort-indexed family of variables $\VAR$, such as presented in \cite{harper_foundations}. For the full definitions we refer to the full version of the paper.

@@

\begin{example}\label{ex:abstract_syntax}
    %To illustrate how the generalized editor calculus could be instantiated for some language, we will specialize an editor calculus as we proceed through this section for the abstract syntax shown in \cref{fig:example_abstract_syntax}.
    To illustrate how the generalized editor calculus could be instantiated for some language, we will through our examples specialize an editor calculus. The abstract syntax for our specialization language is shown in \cref{fig:example_abstract_syntax}.
    \federeboks{R C L L L}{
        \text{Sort} && \text{Term} & \text{Operator} & \text{Arity} \\
        s & ::= & \text{let $x = e$ in $s$} & \text{let} & (e, e.s)s \\
        & | & e & \text{exp} & (e)s \\
        e & ::= & e_1 + e_2 & \text{plus} & (e,e)e \\
        & | & n & \text{num}[n] & ()e \\
        & | & x & \text{var}[x] & ()e
    }{\caption{Abstract syntax of arithmetic expressions and local declarations}\label{fig:example_abstract_syntax}}
    The syntax in \cref{fig:example_abstract_syntax} describes a very simple language of statements and expressions. For example, using the abstract syntax we could construct the statement
    \begin{equation*}
        \text{let $x=5$ in let $y=10$ in $x+y$}
    \end{equation*}
    which, using operators, would be represented as the abt
    \begin{equation*}
        \text{let}(5;x.\text{let}(10;y.\text{exp}(\text{plus}(x;y))))
    \end{equation*}

    Given the abstract syntax in \cref{fig:example_abstract_syntax} we can extend it with holes and cursors. This gives us the new operators shown in \cref{fig:example_holes_and_cursors}.
    \federeboks{R C L L L}{
        \text{Sort} && \text{Term} & \text{Operator} & \text{Arity} \\
        s & ::= & [s] & \text{cursor}_s & (s)s \\
        & | & \hole{s} & \text{hole}_s & ()s \\
        e & ::= & [e] & \text{cursor}_e & (e)e \\
        & | & \hole{e} & \text{hole}_e & ()e
    }{\caption{Introduction of cursors and holes to the abstract syntax in \cref{fig:example_abstract_syntax}}\label{fig:example_holes_and_cursors}}
    With this extension we can now represent statements with holes and cursors as abt's. For example the statement
    \begin{equation*}
        \text{let $x=[5]$ in $x+\hole{e}$}
    \end{equation*}
    would be represented as the abt
    \begin{equation*}
        \text{let}(\text{cursor}_e(5); x.\text{exp}(\text{plus}(x;\text{hole}_e)))    
    \end{equation*}
\end{example}

\subsubsection{Editor Calculus}
The abstract syntax of the generalized editor calculus is shown in \cref{fig:editor_calculus_syntax}. It closely resembles that of the calculus presented by \cite{type_safe_structure_editor}.
\federeboks{R C L}{
    E & ::= & \pi.E \ | \ \phi => E_1|E_2 \ | \ E_1 \ggg E_2 \ | \ \text{rec} \ x.E \ | \ x \ | \ \text{nil} \\
    \pi & ::= & \text{child} \ n \ | \ \text{parent} \ | \ \{o\} \\
    \phi & ::= & \neg\phi \ | \ \phi_1 \wedge \phi_2 \ | \ \phi_1 \vee \phi_2 \ | \ @o \ | \ \Diamond o \ | \ \square o
}{\caption{Abstract syntax of general editor calculus}\label{fig:editor_calculus_syntax}}
Editor expressions $E \in Edt$, where $Edt$ is the sort of editor expressions, describes the behaviour of the editor calculus. The semantics behind the different constructs are defined in section \cref{sec:Semantics}.

Evaluation of abt's, which \cite{type_safe_structure_editor} introduces through a $eval$ construct, is not included in our generalized version. This omission is due to evaluation being entirely dependent on the dynamics of the language in focus, and not the static and structural properties the abstract syntax describes. To define the concept of evaluation in a generalized manner, it would be necessary to devise a formal and concise approach for specifying the semantics of operators.

\begin{example}\label{ex:editor_calculus}
    Expanding upon \cref{ex:abstract_syntax}, we can specialize the abstract syntax of the general editor calculus presented in \cref{fig:editor_calculus_syntax} for our simple language by defining the set of operators $o$ ranges over. In this case we have that
    \begin{equation*}
        o \in \{ \text{let}, \text{exp}, \text{hole}_s, \text{plus}, \text{num}[n], \text{var}[x], \text{hole}_e \}
    \end{equation*}
    allowing us to write editor expressions such as $(@\text{hole}_e => \{\text{plus}\}.\text{nil}|\text{nil})$, which would substitute the current tree encapsulated by the cursor with the $plus$ operator, if the cursor is at the operator $\text{hole}_e$.
\end{example}

\subsubsection{Cursor Context}
In editor expressions, we frequently refer to performing actions on, or with, the abt that is currently encapsulated by the cursor. To support this notion we introduce cursor contexts $C$, inspired by the zipper data structure by \cite{huet_zipper}, as a way to locate the cursor in an abt. Central to this idea is the cursor context $\ctxhole$, or context hole, which specifies the subtree within which the cursor must reside, either as the root, or one of the first children.  For the full definition of cursor contexts, we refer to the full version of the paper, or to \cite{type_safe_structure_editor}, as they have a very similar construct.

@@

\subsubsection{Editor Expressions} \label{sec:Semantics}
The labelled transition system for editor expressions is defined as $(Edt \times \ABT{}, Apc \cup \{\epsilon\}, \Rightarrow)$. Transitions are of the form $\lr{E, a} \ltrans{\alpha} \lr{E', a'}$ where the editor expression $E \in Edt$ is closed and the abt $a \in \ABT{}$ is well-formed. The labels of the transitions, $\alpha$, are the atomic prefix commands $Apc$ and the silent transition $\epsilon$. The transition relation $\Rightarrow$ is defined by the rules shown in \cref{fig:editor_expression_rules}. These facilitate ways of composing editor expressions together with conditions, sequential composition and recursion. The means of traversing and modifying an abt are provided by the prefixed expression $\pi.E$, which evaluates $\pi$ before continuing with $E$. The conditional expression $\phi => E_1|E_2$ reduces to $E_1$ if $\phi$ is satisfied and $E_2$ otherwise. The sequential expression $E_1 \ggg E_2$ evaluates $E_2$ only once $E_1$ has been reduced to $nil$. The recursive expression $rec \ x.E$ binds the recursion variable $x$ in $E$, which can then be used to recursively iterate an expression.

@@

%A concrete overview of the calculus and the constructs necessary to encode every central concept of the editor calculus can be seen in \cref{tab:encoding_requirements}.

%\begin{table}[H]
%    \centering
%    \begin{tabular}{| l | l |}
%        \hline
%        Editor calculus & Encoding\\
%        \hline
%        \ Abstract Binding Trees & \ Simply typed lambda calculus, \stlc \\
%        \hline
%        \ Atomic Prefix Commands & \ \stlc with pattern matching, $\lambda_{->,p}$\\
%        \hline
%        \ Modal Logic & \ \stlc with pattern matching and recursion, $\lambda_{->,p,fix}$ \\
%        \hline
%        \ Editor Expressions & \ \stlc with pattern matching and recursion, $\lambda_{->,p,fix}$ \\
%        \hline 
%    \end{tabular}
%    \vspace{1mm}
%    \caption{Calculus requirements to encode the generalized editor calculus}
%    \label{tab:encoding_requirements}
%\end{table}

@@

>>>>

\documentclass[sigplan,anonymous,review]{acmart}
\input{preamble}

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

\acmConference[GPCE 2023]{Make sure to enter the correct
  conference title from your rights confirmation emai}{July 07--07,
  2023}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A type-safe generalized editor calculus (Short Paper)}

@@

\begin{abstract}
In this paper, we present a generalization of a syntax-directed editor calculus, which can be used to instantiate a specialized syntax-directed editor for any language, given by some abstract syntax. The editor calculus guarantees the absence of syntactical errors while allowing incomplete programs. The generalized editor calculus is then encoded into a simply typed lambda calculus, extended with pairs, booleans, pattern matching and fixed points.
\end{abstract}

@@

\section{Introduction}

Syntax-directed editors aka projectional editors are specialized
editors that allow programmers to create programs directly using the
syntax of the programming language. A prominent early example of an
editor of this kind is that of the Cornell Program Synthesizer by Reps and
Teitelbaum \cite{10.1145/358746.358755}, that was designed with the
PL/I language in mind.

More recent developments introduce models of computation that are
specialized towards this form of editing. The Hazelnut framework by
Omar et al. \cite{hazelnut} and the work by Godiksen et al. and
\cite{type_safe_structure_editor} introduce editor calculi that
describe the syntax and operational semantics of editing commands,
including how such commands can be combined. This line of work has
been tailored specifically towards editing abstract syntax trees from
an applied $\lambda$-calculus.

A major advantage of this approach is that it is typed -- each such editor calculus
is equipped with a type system that guarantees that if an editor term
$E$ is well-typed, then $e$ will produce a well-formed program.

However, existing editor calculi do not account for context-sensitive
aspects of program syntax, and notable not the binding 
mechanisms that are prominenet in abstract syntax.

Moreover, the work so far has been dependent on the syntax of the
programming language that the editor must support. Whenever the
programming language changes, the type safety results for the editor
calculus have had to be proved anew. An example of this is the work by
Andersen et al \cite{10.1145/3587216.3587221} on extending the editor
calculus of \cite{type_safe_structure_editor} with let-declarations.

In later work, Reps and Teitelbaum showed \cite{10.1145/390011.808247}
how one can generalize their original work to build a Synthesizer
Generator that would create a syntax-directed editor for any given
programming language, given a specification of its syntax in the form
of an attribute grammar. Around the same time, Donzeau-Gouge et
al. developed the Mentor programming environment
\cite{10.5555/800054.801990} that involved a command language, Mentol,
and allowed designers to specify language-specific specializations
using a specification language, Metal.

In this paper we propose a generalized editor calculus that deals with
these two issues.

Our calculus deals with binding mechanisms by using notions from
higher-order abstract syntax \cite{hoas}. Moreover, and importantly,
we show how to encode encoded our generalized editor calculus into
versions of the simply-typed lambda calculus extended with pairs,
booleans, pattern matching and fixed-points. These fragments are all
known to have type systems with strong safety guarantees, and in this
way we a general result on type safety that applies to any instance of
our generalized editor calculus.

\section{A Generalized Editor Calculus}\label{sec:general_editor}

We first present a generalization of the syntax-directed editor
calculus by \cite{type_safe_structure_editor}.

\subsection{Representing abstract syntax}

We assume the abstract syntax is given by a set of sorts $\SORT$, an arity-indexed family of operators $\OP$ and a sort-indexed family of variables $\VAR$, such as presented in \cite{harper_foundations}.

@@

\begin{example}\label{ex:abstract_syntax}
Consider the following simple language of expressions with local declarations.
\begin{align*}
  \text{Sort} && \text{Term} & \text{Operator} & \text{Arity} \\
        s & ::= & \text{let $x = e$ in $s$} & \text{let} & (e, e.s)s \\
        & | & e & \text{exp} & (e)s \\
        e & ::= & e_1 + e_2 & \text{plus} & (e,e)e \\
        & | & n & \text{num}[n] & ()e \\
        & | & x & \text{var}[x] & ()e
\end{align*}
 We can here represent
    \begin{equation*}
        \text{let $x=5$ in let $y=10$ in $x+y$}
    \end{equation*}
    as the abt
    \begin{equation*}
        \text{let}(5;x.\text{let}(10;y.\text{exp}(\text{plus}(x;y))))
    \end{equation*} 
%
    We can extend this language with holes and cursors and get
    %
\begin{align*}
        \text{Sort} && \text{Term} & \text{Operator} & \text{Arity} \\
        s & ::= & [s] & \text{cursor}_s & (s)s \\
        & | & \hole{s} & \text{hole}_s & ()s \\
        e & ::= & [e] & \text{cursor}_e & (e)e \\
        & | & \hole{e} & \text{hole}_e & ()e
 \end{align*}
%
    With this extension the statement
    \begin{equation*}
        \text{let $x=[5]$ in $x+\hole{e}$}
    \end{equation*}
    is represented as the abt
    \begin{equation*}
        \text{let}(\text{cursor}_e(5); x.\text{exp}(\text{plus}(x;\text{hole}_e)))    
    \end{equation*}
\end{example}

\subsubsection{The editor calculus}
Editor expressions $E \in \Edt$, where $\Edt$ is the sort of editor
expressions, describe the possible edit actions. The formation rules
are given below.
%
\begin{align*}
  E & ::= & \pi.E  \mid  \phi => E_1|E_2  \mid  E_1 \ggg E_2  \mid  \text{rec} \ x.E  \mid  x  \mid  \text{nil} \\
    \pi & ::= & \text{child} \ n  \mid  \text{parent}  \mid  \{o\} \\
    \phi & ::= & \neg\phi  \mid  \phi_1 \wedge \phi_2  \mid  \phi_1 \vee \phi_2  \mid  @o  \mid  \Diamond o  \mid  \square o
\end{align*}
%
Evaluation of abt's, which \cite{type_safe_structure_editor} introduces through a $eval$ construct, is not included in our generalized version. This omission is due to evaluation being entirely dependent on the dynamics of the language in focus, and not the static and structural properties the abstract syntax describes. To define the concept of evaluation in a generalized manner, it would be necessary to devise a formal and concise approach for specifying the semantics of operators.

\begin{example}\label{ex:editor_calculus}
    We can define a version of the editor calculus for
    \cref{ex:abstract_syntax} by defining the set of operators $o$ ranges over. In this case we have that
    \begin{equation*}
        o \in \{ \text{let}, \text{exp}, \text{hole}_s, \text{plus}, \text{num}[n], \text{var}[x], \text{hole}_e \}
    \end{equation*}
    We can then write e.g. $(@\text{hole}_e =>
    \{\text{plus}\}.\text{nil}|\text{nil})$. This expression would
    substitute the current tree encapsulated by the cursor with the
    $plus$ operator, if the cursor is at the operator
    $\text{hole}_e$. 
\end{example}

\subsubsection{Cursor contexts}
In editor expressions, we frequently refer to performing actions on, or with, the abt that is currently encapsulated by the cursor. To support this notion we introduce cursor contexts $C$, inspired by the zipper data structure by \cite{huet_zipper}, as a way to locate the cursor in an abt. Central to this idea is the cursor context $\ctxhole$, or context hole, which specifies the subtree within which the cursor must reside, either as the root, or one of the first children.  For the full definition of cursor contexts, we refer to the full version of the paper, or to \cite{type_safe_structure_editor}, as they have a very similar construct.

@@

\subsubsection{Editor Expressions} \label{sec:Semantics}

The labelled transition system for editor expressions is defined as
$(\Edt \times \ABT{}, Apc \cup \{\epsilon\},
\Rightarrow)$. Transitions are of the form $\lr{E, a} \ltrans{\alpha}
\lr{E', a'}$ where the editor expression $E \in \Edt$ is closed and
the abt $a \in \ABT{}$ is well-formed. The labels of the transitions,
$\alpha$, are the atomic prefix commands $Apc$ and the silent
transition $\epsilon$. The transition relation $\Rightarrow$ is
defined by the rules shown in
\cref{fig:editor_expression_rules}. These facilitate ways of composing
editor expressions together with conditions, sequential composition
and recursion. The means of traversing and modifying an abt are
provided by the prefixed expression $\pi.E$, which evaluates $\pi$
before continuing with $E$. The conditional expression $\phi =>
E_1|E_2$ reduces to $E_1$ if $\phi$ is satisfied and $E_2$
otherwise. The sequential expression $E_1 \ggg E_2$ evaluates $E_2$
only once $E_1$ has been reduced to $nil$. The recursive expression
$rec \ x.E$ binds the recursion variable $x$ in $E$, which can then be
used to recursively iterate an expression.

@@

A concrete overview of the calculus and the constructs necessary to encode every central concept of the editor calculus can be seen in \cref{tab:encoding_requirements}.

\begin{table*}
   \centering
   \begin{tabular}{| l | l |}
       \hline
       \textsf{Editor calculus} & \textsf{Encoding}\\
       \hline\hline
       \ Abstract Binding Trees & \ Simply typed lambda calculus, \stlc \\
       \hline
       \ Atomic Prefix Commands & \ \stlc with pattern matching, $\lambda_{->,p}$\\
       \hline
       \ Modal Logic & \ \stlc with pattern matching and recursion, $\lambda_{->,p,fix}$ \\
       \hline
       \ Editor Expressions & \ \stlc with pattern matching and recursion, $\lambda_{->,p,fix}$ \\
       \hline 
   \end{tabular}
   \vspace{1mm}
   \caption{Editor calculus fragments and the lambda calculus fragments
     used for encoding them}
   \label{tab:encoding_requirements}
\end{table*}

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
